---
title: "CNV analysis"
output: html_notebook
---

```{r}
library(tidyverse)
library(SCpubr)
```

# Load infercnvpy output
```{r}
cnv_calls <- read.delim(gzfile("~/Projects/tnbc_cna/output/infercnvpy_matrix.txt.gz"),
                        sep="\t", header=F)
cnv_calls <- as.matrix(cnv_calls)
#rows = cells, cols = genomic bins
```

# Calculate % Genome Altered
```{r}
meta <- read.csv("~/Projects/tnbc_cna/output/infercnvpy_cell_metadata.csv",
                 row.names=1)
df_nonmalignant <- cnv_calls %>%
  as.data.frame() %>%
  mutate(celltype_pred = meta$celltype_pred) %>%
  filter(celltype_pred != "Epithelial") %>% # remove malignant cells
  filter(celltype_pred != "Mast") %>% #remove mast.  artifactually high cna inference
  pivot_longer(-celltype_pred, names_to="Bin", values_to="Value") %>%
  group_by(Bin) %>%
  summarize(mean = mean(Value), sd = sd(Value)) %>%
  mutate(upper_thresh = mean + 2*sd, #Threshold = mean +/- 2SD
         lower_thresh = mean - 2*sd)


df_malignant <- cnv_calls %>%
  as.data.frame() %>%
  mutate(leiden = meta$leiden,
         cell_id = rownames(meta)) %>%
  filter(leiden %in% c(0, 6)) %>%
  column_to_rownames(var="cell_id")
df_malignant$celltype_pred <- NULL
df_malignant <- as.matrix(df_malignant)

df_malignant <- df_malignant[,df_nonmalignant$Bin]

altered <- list()
for(i in 1:ncol(df_malignant)){
  altered[[i]] <- (df_malignant[,i] > df_nonmalignant$upper_thresh[i]) |
    (df_malignant[,i] < df_nonmalignant$lower_thresh[i])
}

altered <- do.call(cbind, altered)

fga <- rowSums(altered) / ncol(altered)
```

Export data
```{r}
#export for khalid
pam50 <- read.csv("../output/pam50_subtype_annotations.csv", row.names=1)
fga_df <- data.frame(cell = names(fga), 
                     fga = fga,
                     sample_id = meta[names(fga), "sample_id"],
                     subtype = meta[names(fga), "subtype"])
fga_df$pam50 <- pam50$pam50_subtype[match(fga_df$sample_id, rownames(pam50))]

# summarize by patient
fga_df <- fga_df %>%
  group_by(sample_id, subtype, pam50) %>%
  summarize(fga = mean(fga))
write.csv(fga_df, file="../output/fga_summary.csv", row.names = F)
```

```{r}
fga_df$tmp <- ifelse(fga_df$pam50 == "BLBC", "Yes", "No")
summary(lm(fga_df$fga ~ fga_df$tmp))
```


# Aggregate by chromosome arm
```{r}
arm_locs <- human_chr_locations #loaded with package. based on GRCh38, just like locations in var

arms <- arm_locs %>% filter(arm=='q') #has proper end
arms <- arms %>% filter( chr != "Y")
arms$rel_p_length <- arms$start / arms$end

#This is based on the specific number of bins per chromosome from the infercnvpy output
chr_bins <- data.frame(
  chr = c(paste0("chr", 1:22), "chrX"),
  start = c(0, 152, 252, 337, 392, 459, 533, 603, 654, 712, 769, 854, 933, 960,
            1011, 1057, 1123, 1210, 1211, 1323, 1363, 1364, 1401)
)

chr_bins$start <- chr_bins$start + 1 #adjust 0-based to 1-based count
chr_bins$end <- c((chr_bins$start[2:nrow(chr_bins)]-1), 1451)

chr_bins$length <- c(chr_bins$start[2:nrow(chr_bins)], 1451) - chr_bins$start

chr_bins$q_start <- chr_bins$start + ((chr_bins$end - chr_bins$start) * arms$rel_p_length)
chr_bins$q_start <- round(chr_bins$q_start)

head(chr_bins)
```


```{r}
chr_order <- rep(chr_bins$chr, each = 2)
chr_order <- paste0(chr_order, c("_p", "_q"))

cnv_list <- list()
#Get average CNV signal for all bins within arm range
for(i in 1:nrow(chr_bins)){
  print(chr_bins$chr[i])
  if(chr_bins$length[i] > 1){
    p_vals <- rowMeans(cnv_calls[,chr_bins$start[i]:chr_bins$q_start[i]])
    q_vals <- rowMeans(cnv_calls[,chr_bins$q_start[i]:chr_bins$end[i]])
  } else{
    p_vals <- cnv_calls[,chr_bins$start[i]]
    q_vals <- cnv_calls[,chr_bins$start[i]]
  }
  
  cnv_list[[i]] <- data.frame(p = p_vals, q=q_vals)
  colnames(cnv_list[[i]]) <- paste0(chr_bins$chr[i], c("p", "q"))
}

cnv_list <- do.call('cbind', cnv_list)
cnv_list <- as.matrix(cnv_list)
cnv_list[cnv_list > 0.1] <- 0.1
cnv_list[cnv_list < -0.1] <- -0.1
```

```{r}
#Arm-level quantifications for all cells in the dataset
write.table(cnv_list, file="~/Projects/tnbc_cna/output/infercnvpy_matrix_arm_binned.csv", 
          col.names = F, row.names = F, quote = F, sep=",")
```

# Load cell metadata
```{r}
meta <- read.csv("~/Projects/tnbc_cna/output/infercnvpy_cell_metadata.csv",
                 row.names=1)
#We'll also add pam50 annotations, which were present when we ran inferCNVpy
pam50_anno <- read.csv("~/Projects/tnbc_cna/output/pam50_subtype_annotations.csv",
                       row.names=1)
meta$pam50_subtype <- pam50_anno$pam50_subtype[match(meta$sample_id, rownames(pam50_anno))]
#write.csv(meta, file="~/Projects/tnbc_cna/output/infercnvpy_cell_metadata_withPAM50.csv",
#          row.names=F)

cnv_list <- read.csv("~/Projects/tnbc_cna/output/infercnvpy_matrix_arm_binned.csv",
                     header = F)
colnames(cnv_list) <- paste0(rep(chr_bins$chr, each=2), "_", c("p", "q"))
rownames(cnv_list) <- rownames(meta)
```

```{r}
table(rownames(cnv_list) == rownames(meta))
head(cnv_list)
```

# CNV calls
We'll define thresholds for each arm based on the signal in non-malignant cells
For each arm, first calculate the average for each cell type, then take average of all of those. This ensures it is not as affected by cell count per cell type

 
```{r}
nonmalignant_values <- cnv_list %>%
  mutate(celltype_pred = meta$celltype_pred, # Add cell type information
         sample_id = meta$sample_id) %>%
  pivot_longer(-c(celltype_pred, sample_id), names_to="Arm", #Tidy
               values_to="CNA") %>%
  filter(celltype_pred != "Epithelial") %>% # remove malignant cells
  filter(celltype_pred != "Mast") %>% #remove mast. seems to have artifactually high cna inference
  group_by(celltype_pred, Arm) %>%
  summarize(CNA_mean = mean(CNA), CNA_SD = sd(CNA)) %>% # mean and sd of CNA signal per cell type
  ungroup() %>% group_by(Arm) %>%
  summarize(CNA_SD = mean(CNA_SD), CNA_mean = mean(CNA_mean)) %>% #average of those mean and SD values
  mutate(upper_thresh = CNA_mean + 1*CNA_SD, #Threshold = mean +/- 2SD
         lower_thresh = CNA_mean - 1*CNA_SD)


# Match order w/ cnv_list
nonmalignant_values <- nonmalignant_values[match(colnames(cnv_list), nonmalignant_values$Arm),]
```

## Per cell
```{r}
cnv_calls <- list()
for(i in 1:ncol(cnv_list)){
  values = cnv_list[,i]
  calls <- rep(0, length(values))
  calls[values > nonmalignant_values$upper_thresh[i]] <- 1
  calls[values < nonmalignant_values$lower_thresh[i]] <- -1
  cnv_calls[[i]] <- calls
}

cnv_calls <- do.call('cbind', cnv_calls)
rownames(cnv_calls) <- rownames(cnv_list)
colnames(cnv_calls) <- colnames(cnv_list)
```

# Clonality of mutations
Subset to epithelial only, group by patient, count proportion of cells w/ alteration in each arm
```{r}
clonality <- cnv_calls %>%
  as.data.frame() %>%
  mutate(sample_id = meta$sample_id, 
         pam50 = meta$pam50_subtype,
         celltype_pred = meta$celltype_pred) %>%
  filter(celltype_pred == "Epithelial") %>%
  pivot_longer(-c("sample_id", "pam50", "celltype_pred"), 
               names_to = "Arm", values_to = "CNA") %>%
  group_by(sample_id, pam50, Arm) %>%
  summarize(Gain_sum = sum(CNA==1),
            Loss_sum = sum(CNA == -1),
            Gain_prop = Gain_sum / n(),
            Loss_prop = Loss_sum / n(),
            Cell_count = n())

write.csv(clonality, file="../output/cnv_clonality.csv", row.names=F)
clonality <- read.csv("../output/cnv_clonality.csv")
head(clonality) # Per sample, number of cells with alteration (gain or loss) in a given arm
```

Make it so each Arm entry specifies gain or loss (eg. chr10_p_gain)
```{r}
chr_remove <- c("chrX_p", "chrX_q", "chr13_p", 
                "chr14_p", "chr15_p", "chr21_p", "chr22_p")
clonality <- filter(clonality, Arm %nin% chr_remove)

df_gain <- clonality
df_gain$Arm <- paste0(df_gain$Arm, "_gain")
df_gain$Prop <- df_gain$Gain_prop

df_loss <- clonality
df_loss$Arm <- paste0(df_loss$Arm, "_loss")
df_loss$Prop <- df_loss$Loss_prop

clonality <- rbind(df_gain, df_loss)
```


Plot clonality of alterations
```{r}
cols <- RColorBrewer::brewer.pal(9, "BuPu")
```

##Plot
```{r}
plotClonality <- function(subtype){
  df <- clonality %>% filter(pam50 == subtype)
  
  tumor_count <- length(unique(df$sample_id))
  
  #Count alterations
  df$clonality <- "Not altered"
  
  df$clonality[df$Prop > 0.1 & df$Prop < 0.5] <- "Subclonal"
  df$clonality[df$Prop > 0.5] <- "Clonal"
  
  # Get arm ranking
  counts <- table(df$Arm, df$clonality)
  chr_order <- rownames(counts)[order(counts[,"Clonal"])]
  df$Arm  <- factor(df$Arm, levels = rev(chr_order))
  
  df$clonality <- factor(df$clonality, levels = c("Not altered", "Subclonal", "Clonal"))
  
  p <- ggplot(df, aes(x=Arm, fill=clonality)) +
    geom_bar(position="fill") +
    ggtitle(paste0("Arm alteration: ", 
                   subtype, " (", tumor_count, " patients)")) +
    scale_y_continuous(expand=c(0,0)) +
    scale_fill_manual(values=c(cols[2], cols[4], cols[8]),
                      name="") +
    xlab("") + ylab("Proportion of samples") +
    theme_classic() +
    theme(axis.text.x = element_text(angle=45, hjust=1, 
                                     color="black", size=6),
          axis.text.y = element_text(color="black"))
  return(p)
}
```

```{r}
ggsave(plotClonality("BLBC"),
       filename="../figs/CNBA_clonality_BLBC.pdf",
       width=12, height=3)

ggsave(plotClonality("ER.A"),
       filename="../figs/CNBA_clonality_ERA.pdf",
       width=12, height=3)

ggsave(plotClonality("ER.B"),
       filename="../figs/CNBA_clonality_ERB.pdf",
       width=12, height=3)

ggsave(plotClonality("HER2"),
       filename="../figs/CNBA_clonality_HER2.pdf",
       width=12, height=3)

ggsave(plotClonality("Normal.CL"),
       filename="../figs/CNBA_clonality_NormalCL.pdf",
       width=12, height=3)
```


## Previous version split by gains and losses
```{r}
plotClonality <- function(subtype, alteration){
  df <- clonality %>% filter(pam50 == subtype)
  
  tumor_count <- length(unique(df$sample_id))
  
  #Count alterations
  df$clonality <- "Not altered"
  
  if(alteration == "Gain"){
    df$clonality[df$Gain_prop > 0.1 & df$Gain_prop < 0.5] <- "Subclonal"
    df$clonality[df$Gain_prop > 0.5] <- "Clonal"
  } else if(alteration == "Loss"){
    df$clonality[df$Loss_prop > 0.1 & df$Loss_prop < 0.5] <- "Subclonal"
    df$clonality[df$Loss_prop > 0.5] <- "Clonal"
  }
  
  # Get arm ranking
  counts <- table(df$Arm, df$clonality)
  chr_order <- rownames(counts)[order(counts[,"Clonal"])]
  df$Arm  <- factor(df$Arm, levels = rev(chr_order))
  
  df$clonality <- factor(df$clonality, levels = c("Not altered", "Subclonal", "Clonal"))
  
  p <- ggplot(df, aes(x=Arm, fill=clonality)) +
    geom_bar(position="fill") +
    ggtitle(paste0("Arm ", alteration, ": ", 
                   subtype, " (", tumor_count, " patients)")) +
    scale_y_continuous(expand=c(0,0)) +
    scale_fill_manual(values=c(cols[2], cols[4], cols[8]),
                      name="") +
    xlab("") + ylab("Proportion of samples") +
    theme_classic() +
    theme(axis.text.x = element_text(angle=45, hjust=1, color="black"),
          axis.text.y = element_text(color="black"))
  return(p)
}
```

```{r}
p1 <- plotClonality("BLBC", "Gain")
p2 <- plotClonality("BLBC", "Loss")
clonality_plot <- cowplot::plot_grid(p1, p2,  ncol=1, align='v')
cowplot::save_plot(clonality_plot,
                   filename="../figs/CNA_clonality_BLBC.pdf",
                   base_width=10, base_height=4.5)

p1 <- plotClonality("HER2", "Gain")
p2 <- plotClonality("HER2", "Loss")
clonality_plot <- cowplot::plot_grid(p1, p2,  ncol=1, align='v')
cowplot::save_plot(clonality_plot,
                   filename="../figs/CNA_clonality_HER2.pdf",
                   base_width=10, base_height=4.5)

p1 <- plotClonality("ER.A", "Gain")
p2 <- plotClonality("ER.A", "Loss")
clonality_plot <- cowplot::plot_grid(p1, p2,  ncol=1, align='v')
cowplot::save_plot(clonality_plot,
                   filename="../figs/CNA_clonality_ER.A.pdf",
                   base_width=10, base_height=4.5)

p1 <- plotClonality("ER.B", "Gain")
p2 <- plotClonality("ER.B", "Loss")
clonality_plot <- cowplot::plot_grid(p1, p2,  ncol=1, align='v')
cowplot::save_plot(clonality_plot,
                   filename="../figs/CNA_clonality_ER.B.pdf",
                   base_width=10, base_height=4.5)

p1 <- plotClonality("Normal.CL", "Gain")
p2 <- plotClonality("Normal.CL", "Loss")
clonality_plot <- cowplot::plot_grid(p1, p2,  ncol=1, align='v')
cowplot::save_plot(clonality_plot,
                   filename="../figs/CNA_clonality_Normal.CL.pdf",
                   base_width=10, base_height=4.5)
```

# Mean-variance clonality
The relationship between the arm's mean CNA call and variance should tell us about clonality
High mean, low variance = clonal
high mean, high variance = subclone
low mean, low variance = not altered

Will first try just summarizing this at the level of subtypes
```{r}
mean_var <- cnv_list %>%
  mutate(celltype_pred = meta$celltype_pred, # Add cell type information
         sample_id = meta$sample_id,
         pam50 = meta$pam50_subtype) %>%
  filter(celltype_pred == "Epithelial") %>%
  pivot_longer(-c(celltype_pred, sample_id, pam50), names_to="Arm", #Tidy
               values_to="CNA") %>%
  group_by(pam50, sample_id, Arm) %>%
  summarize(CNA_mean = mean(CNA), CNA_var = var(CNA)) %>%
  ungroup() %>% group_by(pam50, Arm) %>%
  summarize(CNA_mean = mean(CNA_mean), CNA_var = mean(CNA_var))
```

```{r}
var_plot <- ggplot(mean_var %>% filter(pam50 == "BLBC"), 
                   aes(x=CNA_mean, y=CNA_var, label=Arm)) +
  geom_point() +
  geom_label_repel(size = 2) +
  theme_classic()
```





